The Rockerverse: Packages and Applications for Containerization with R

### 2020-01-27 manuscript.tex
update tex file

% !TeX root = RJwrapper.tex
\title{The Rockerverse: Images, Packages, and Applications for Containerization
with R}
\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao, Charlotte Van Petegem}

@@

\abstract{%
The Rocker~Project provides widely-used Docker images for R across
different application scenarios. This articles surveys downstream
projects building upon Rocker and presents the current state of R
packages for managing Docker images and controlling containers. We also
look beyond Rocker to other projects connecting containerisation with R,
namely alternative suites of images. These use cases and the variety of
applications demonstrate the power of Rocker specifically and
containerisation in general. We identified common themes across this
diversity: reproducible environments, scalability and efficiency, and
portability across clouds.
}

@@

The article continues with a brief introduction of containerization
basics and the Rocker~Project, followed by descriptions of projects
developing images with R installations. Then, we present applications,
starting with the R packages specificly for interacting with Docker,
second-level packages using containers indirectly or only for specific
features, up to complex use cases leveraging containers. We conclude
with a reflection on the landscape of packages and applications and
point out future directions of development.

@@

For statistical computing and analysis centered around R, the
Rocker~Project has provided a variety of Docker containers since its
start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides several
lines of containers spanning to from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for better maintainability (i.e.,
smaller \texttt{Dockerfile}s), composability, and to reduce build time.
Also of note is a series of ``versioned'' containers which match the R
release they contain with the \emph{then-current} set of packages via
the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, who provide findung for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

\hypertarget{container-images}{%
\section{Container images}\label{container-images}}

\label{images}

\hypertarget{images-for-alternative-r-engines}{%
\subsection{Images for alternative R
engines}\label{images-for-alternative-r-engines}}

\label{alternatives}

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled
variety of established and upcoming features. Nevertheless, special use
cases and experimental projects in academia and industry exist to test
new approaches, or provide features different to what the R
implementation maintained by the
\href{https://www.r-project.org/contributors.html}{R Core Team} and
available via \href{https://cran.r-project.org/}{CRAN} provides. While
the images presented in this section are far from being as vetted,
stable, and widely used as any of the Rocker images, they demonstrate an
important advantage of containerisation technology, namely the abiliy to
transparently build portable stacks of open source software and make
them easily accessible to others. All these alternative implementations
of R are published under GPL licenses. Since all implementations of R
claim better performance as a core motivation, a comparision based on
Docker images, potentially leveraging the
\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
restriction mechanisms} of Docker to level the playing field, seems
useful future work.

\textbf{Microsoft R Open} (MRO) is an R distribution formerly known as
Revolution R Open (RRO) before Revolution Analytics was acquired by
Microsoft. MRO is compatible with main R and its packages and it
``{[}..{]} includes additional capabilities for improved performance,
reproducibility, and platform support.'' \citep{microsoft_mro} Most
notably these capabilities are the MRAN repository, which is enabled by
default, and some bundled packages such as \CRANpkg{checkpoint} and
specific R packages maintained by
Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
Originally the optional integration with
\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
(MKL) for multi-threading in linear algebra operations
\citep{microsoft_multithread} was a distinguishing feature before it was
made available publicly, and can be used in base R, too
\citep{eddelbuettel_thinking_2018}.

MRO does not provide official Docker images, but a set of
community-maintained \texttt{Dockerfile}s and Docker images are provided
by the poject \texttt{mro-docker}
(\url{https://github.com/nuest/mro-docker}), e.g., the image
\href{https://hub.docker.com/repository/docker/nuest/mro/}{\texttt{nuest/mro}}.
The images are inspired by the Rocker images and can be used much in the
same fashion, effectively a drop-in replacement allowing users to
quickly evaluate if the benefits of MRO + Intel® MKL apply to their use
case. Version-tagged images are provided for the latest bugfix release
of recent R versions. Extended license information about MKL is printed
at every startup.

\textbf{Renjin} is an interpreter for the R language running on top of
the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
Virtual Machine} (JVM) providing full two-way access between Java and R
code \citep{wikipedia_renjin_2018}. It was developed to combine the
benefits of R, such scripting and extension packages, with the JVM's
advantages in the areas of security, cross-platform availability and
integration into enterprise platforms. R extension packages need to be
specially compiled and are distributed via the Java package manager
\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf.~\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e., at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g., \texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained
images for selected releases only are available under
\texttt{nuest/renjin} on Docker~Hub and GitHub at
\url{https://hub.docker.com/r/nuest/renjin} and
\url{https://github.com/nuest/renjin-docker} respectively. These images
expose the command line interface of Renjin in a similar fashion as
Rocker images and allow an easy evaluation of Renjin's suitability, but
are not intended for production use.

\textbf{pqR} (\url{http://www.pqr-project.org/}) is a ``a pretty quick
version of R''. \emph{pqR} attempts to improve R on some opinionated
issues in the R language and is the basis for experimental features,
e.g., automatic
diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development
\href{https://github.com/radfordneal/pqR/}{on GitHub} is a one man
project and it does not provide any Docker images. But especially
disruptive approaches may contribute to the development of the R
ecosystem, so the \texttt{nuest/pqr} image was independently created,
see Docker Hub at \url{https://hub.docker.com/r/nuest/pqr/} and GitHub
at \url{https://github.com/nuest/pqr-docker}.

\textbf{FastR} (\url{https://github.com/oracle/fastr}) is ``A
high-performance implementation of the R programming language, built on
GraalVM'' \citep{oracle_labs_oraclefastr_2020}. It is developed by
Oracle, connects R to the GraalVM ecosystem
\citep{wikipedia_graalvm_2019}, and also claims superior performance but
also targets full compatibility with base R
\citep{oracle_labs_oraclefastr_2020}. There are no offical Docker images
provided, but indepedently maintained experimental images and
\texttt{Dockerfile}s are provided by
\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker}, e.g.,
\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
on Docker Hub}.

\hypertarget{bioconductor}{%
\subsection{Bioconductor}\label{bioconductor}}

\label{bioc}

\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1823 R
software packages as of October 30th 2019, as well as packages
containing annotation or experiment data. \emph{Bioconductor} has a
semi-annual release cycle, each release is associated with a particular
version of R, and Docker images are provided for current and past
versions of \emph{Bioconductor} for convenience and reproducibility. All
images, included are described on the \emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), created with
\texttt{Dockerfile}s maintained on GitHub, and distributed through
Docker~Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
\emph{Bioconductor}'s `base' Docker images are built on top of the
\texttt{rocker/rstudio} image. \emph{Bioconductor} installs packages
based on the R version in combination with the Bioconductor version, and
therefore uses Bioconductor version tagging \texttt{devel} and
\texttt{RELEASE\_X\_Y}, e.g., \texttt{RELEASE\_3\_10}. Past and current
combinations of R and \emph{Bioconductor} will therefore be accessible
via a specific image tags. The \emph{Bioconductor} \texttt{Dockerfile}
selects the desired version of R from Rocker, adds required system
dependencies, and uses the \CRANpkg{BiocManager} package for installing
appropriate versions of \emph{Bioconductor} packages. A strength of this
approach is that the responsibility for complex software configuration
(including customized development) is shifted from the user to the
experienced \emph{Bioconductor} core team. A recent audit of the
\emph{Bioconductor} image stack \texttt{Dockerfile} led to the
deprecation of several community-maintained images, because the numerous
specific images became too hard to understand, complex to maintain, and
cumbersome to extent. As part of the simplification, a recent innovation
is to produce a \texttt{bioconductor\_docker:devel} image to emulate the
\emph{Bioconductor} nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the
\emph{system dependencies} needed to install and check almost all (1813
out of 1823) \emph{Bioconductor} software packages and saves users and
package developers from managing these themselves. Furthermore the image
is configured so that \texttt{.libPaths()} has
\texttt{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can persist
installed packages across Docker sessions or image updates. Many R users
pursue flexible work flows tailored to particular analysis needs, rather
than standardized work flows. The new \texttt{bioconductor\_docker}
image is well-suited for this pattern, while
\texttt{bioconductor\_docker:devel} provides developers with a test
environment close to \emph{Bioconductor}'s build system.

\hypertarget{images-for-historic-r-versions}{%
\subsection{Images for (historic) R
versions}\label{images-for-historic-r-versions}}

\label{versions}

As with any other software, each new version of R comes with its share
of changes. Some are breaking changes, some are not, but the fact is
that running a piece of code in a given version might give different
results from another version can be an issue when it comes to
reproducibility of workflows and stability of applications. For example,
think about R random seed: starting with \texttt{R\ 3.6.0}, running
\texttt{set.seed(2811);\ sample(1:1000,\ 2)} will not give the same
result as if it was run inside an older version of R. That might seem
trivial, but all the code using the random number generator will not be
exactly reproducible after this breaking change. Therefore, controlling
the version of software is most crucial for reproducible research
\citep[e.g.][]{boettiger_introduction_2015}.

Containers are perfectly suited to capture a specific configuration of a
computing environment to prohibit such problems. The Rocker images
provides the \emph{versioned stack} for different the version of R since
the project inception, which use the \emph{then-current} stable Debian
base image \citep[cf.][]{RJ-2017-065}. Based on custom build phase
hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}},
i.e., small shell scripts executed at different phases in the automated
build of Rocker images, there are also semantic version tags for the
most recent freezed (i.e., using MRAN) versions with only the major and
minor
version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
So, at the time of writing this article, \texttt{rocker/r-ver:3} and
\texttt{3.6} are aliases for \texttt{3.6.0}, because \texttt{3.6.1} is
the latest release. With the release of \texttt{3.6.2} and pinning of
the MRAN version in \texttt{3.6.1}, the tags \texttt{3} and \texttt{3.6}
will deliver the same image as \texttt{3.6.1}. These tags allow users to
update the base image to retrieve bugfixes while reducing the risk of
also introducing breaking changes.

\href{https://github.com/ColinFay/ronline}{r-online} is an app for
helping users to detect breaking changes between different R versions,
and for historic exploration of R. With a standalone NodeJS app or
\href{https://srv.colinfay.me/r-online}{online}, the user can compare a
piece of code run in two separate versions of R. Internally, r-online
opens one or two Docker instances with the given version of R based on
Rocker images, executes a given piece of code, and returns the result to
the user.

A container also provides an isolated sandbox environment suitable for
testing and evaluation, without interfering with the ``main'' working
environment. This enables cross version testing and bugfixing. To take
this even further, the Rocker contributors are discussing the provision
of R versions reaching back futher than the project's own inception,
reaching back as far as R \texttt{2.x} and \texttt{1.x}
\footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
The main challenges are finding a suitable base image with a matching OS
version, or making the adjustments to compile R on more recent OS
releases.

\hypertarget{windows-images}{%
\subsection{Windows Images}\label{windows-images}}

\label{windows}

Docker containers on the Windows operating system were originally quite
cumbersome to use. They required an extra tool,
\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
only exists as a legacy solution for older Windows systems. Docker
toolbox leverages \texttt{docker-machine} (see also
Section~\nameref{interfaces} for an R package interfacing with
\texttt{docker-machine}) to handle the process of creating a local
virtual host which could host Docker Engine, while exposing the regular
Docker CLI. The Docker CLI commands are forwarded to the virtual host
transparently for the user.

For current Windows Server (2016 and later) and Windows Desktop
(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
requires Windows 10) versions, Docker is supported
natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
and different base images are offered by
Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
The Docker CLI can be used in just the same way as on other operating
systems but not every base image is supported on every Windows
host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
On Docker Desktop for Windows, the user can run both Linux-based and
Windows-based containers, but only one of the Docker daemons can be used
at a
time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.

\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker~Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. All these images can be run on Docker for Desktop on
Windows 10. New R versions are only added on demand. The explorative
rocker-win project demonstrates the variety of Windows on Docker, which
is potentially confusing, e.g., because of the different base images.
The proof of concept provides a starting point for any
Windows-constrained user to leverage containerisation, including
examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R
packages with system dependencies such as \CRANpkg{sf}. The latter is
quite comfortable and fast actually, compared to the default
installation from source on Linux, because CRAN ships pre-compiled
binaries for Windows. However, compared to the various Linux images, the
footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the
development and Docker and Microsoft collaborate
closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about
reproducibility, will prefer the more widely and host independent
Linux-based containers. Naturally, a Windows license is required for the
host machine, and the licensing impacts the potential to build images in
clouds or redistribute exported images as archive files. But the
developments in the \texttt{rocker-win} prototype show that containers
can also be used for R workflows depending on Windows-only tools, for
leveraging an existing Windows Server-based infrastructure where
policies can otherwise not be met, or for streamlining interactions with
system operating staff that runs Windows-based containers themselves.

\hypertarget{non-rocker-linux-images}{%
\subsection{Non-Rocker Linux images}\label{non-rocker-linux-images}}

\label{nondebian}

\label{rhub} The \textbf{R-hub} project provides \emph{``a collection of
services to help R package development''}, with the package builder as
the most prominent one\citep{r-hub_docs_2019}. The builder allows R
package developers to check their R package on different platforms and R
versions using a web form or the package \CRANpkg{rhub}
\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
these checks, taking advantages of their well-defined environments and
sandboxing. The \texttt{Dockerfile}s and images are published on
Docker~Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
and
GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
respectively. The images comprise release and development builds of R
running in the base images of the
\href{https://www.debian.org/}{Debian},
\href{https://ubuntu.com/}{Ubuntu},
\href{https://getfedora.org/}{Fedora}, and
\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
under
development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
see \texttt{rhub::local\_check\_linux\_images()} for a full list), and
are intended for debugging (see Section~\nameref{development}) not for
reproducible workflows. \emph{The rationale} of not re-using the Rocker
image stack is the high number of operatings systems and configurations,
which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the
operating systems, e.g., for \texttt{debian}, adding GCC in
\texttt{debian-gcc} and then R's development version in
\texttt{debian-gcc-devel}. The platform also has a specific user
configuration differing from the setup covering typical applications of
Rocker images, and the images include considerably more software, such
as a fairly complete LaTeX installation, to provide an experience closer
to CRAN.

\href{https://rstudio.com/}{\textbf{RStudio}} also maintains a set of
\href{https://github.com/rstudio/r-docker}{Docker images} for multiple
operating systems including SUSE, Ubuntu, CentOS, and Debian. The image
stack includes an \href{https://github.com/rstudio/r-builds}{opinionated
R installation} and ensures the R installation and profile is consistent
across the different Linux distributions. The base image also installs R
in a versioned directory and uses a minimal set of build and runtime
dependencies. Furthermore these images are responsible for creating
pre-compiled binary R packages for Linux, which supplement the binaries
built by CRAN. These pre-compiled packages dramatically decrease the
installation time of R packages on Linux and subsequently can decrease
the build time of Docker images \citep{lopp_package_2019}.

There are also independent projects installing R or RStudio on
\href{https://www.alpinelinux.org/}{\textbf{Alpine Linux}}, but not
beyond the proof of concept
stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic
and thereby small and secure containers, e.g., in hardware with limited
storage. R-hub's proof of concept
\href{https://github.com/r-hub/r-minimal}{\texttt{r-minimal}} takes this
to the extreme with compressed image size of 20MB and bespoke install
scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are
needed, e.g., if adding compilers, packages, or especially if installing
powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the
\href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
the \texttt{glibc} used in Debian/Ubuntu, for which more tooling and
experiences
exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's `slim' variant and removing capabilities, one
could probalby achieve similarly small image sizes on a more common
stack.

\textbf{\pkg{altRnative}}
(\url{https://github.com/ismailsunni/altRnative/}) is an experimental R
package for running the same code across multiple containerised versions
of R, intended for comparison across operating systems and different
implementations of R. It comes with a collection of \texttt{Dockerfile}s
and corresponding images in multiple combinations, currently including,
e.g., MRO, FastR, Fedora, and
TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.

\hypertarget{data-science-images}{%
\subsection{Data Science images}\label{data-science-images}}

\label{datascience}

\emph{Data Science} is a widely discussed topic among all academic
disciplines \citep[e.g.,][]{donoho_50_2017}. The discussions shed a
light on the tools and craftspersonship behind the analysis of data with
computational methods. The practice of Data Science often involves a
combination tools and software stacks and requires a cross-cutting
skillset. This complexity and an inherent concern for openness and
reproducibility in the Data Science community lead to Docker being used
widely. This section presents exemplary Docker images and image stacks
featuring R intended for Data Science.

The \href{https://github.com/jupyter/docker-stacks/}{\textbf{Jupyter
Docker Stacks}} project are a set of ready-to-run Docker images
containing Jupyter applications and interactive computing tools
\citep{project_jupyter_jupyter_2018}. The \texttt{jupyter/r-notebook}
image includes R and ``popular packages'', and naturally also the
IRKernel (\url{https://irkernel.github.io/}), an R kernel for Jupyter so
that Jupyter Notebooks can contain R code cells. R is also included in
the catchall \texttt{jupyter/datascience-notebook}
image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
For example, these images allow users to quickly start a Jupyter
Notebook server locally or build their own specialised images on top of
stable toolsets. R is installed using the Conda package manager, which
can manage environments for various programming languages, pinning both
the R version and the versions of R
packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.

\textbf{Kaggle} provides the
\href{https://hub.docker.com/r/kaggle/rstats}{\texttt{gcr.io/kaggle-images/rstats}}
image (previously \texttt{kaggle/rstats}) and
\href{https://github.com/Kaggle/docker-rstats}{corresponding
\texttt{Dockerfile}} for usage in their Machine Learning competitions
and easy access to the associated datasets. It includes machine learning
libraries such as Tensorflow and Keras (see also image
\texttt{rocker/ml} in Section~\nameref{rocker-gpu}), and also configures
the \CRANpkg{reticulate} package. The image uses a base image with
\emph{all packages from CRAN}, \texttt{gcr.io/kaggle-images/rcran},
which requires a Google Cloud Build as Docker~Hub would time
out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if
having everything available is actually convenient.

As a further example,
\href{https://radiant-rstats.github.io/docs/}{\textbf{Radiant} project}
provides several images, e.g.,
\href{https://hub.docker.com/r/vnijs/rsm-msba-spark}{\texttt{vnijs/rsm-msba-spark}},
for their browser-based business analytics interface based on
\CRANpkg{Shiny} (\texttt{Dockerfile}
\href{https://github.com/radiant-rstats/docker}{on GitHub}) and for use
in education as part of an MSc course. As data science often applies a
multitude of tools, this image favours inclusion over selection and
features Python, Postgres, JupyterLab and Visual Studio Code besides R
and RStudio, bringing the image size up to 9GB.

\textbf{Gigantum} (\url{http://gigantum.com/} is a platform for open and
decentralized data science with a focus on using automation and
user-friendly tools for easy sharing of reproducible computational
workflows. Gigantum builds on the \emph{Gigantum Client} (running either
locally or on a remote server) for development and execution of
data-focused \emph{Projects}, which can be stored and shared via the
\emph{Gigantum Hub} or via a zipfile export. The Client is a user
friendly interface to a backend using Docker containers to package,
build, and run Gigantum projects. It is configured to use a default set
of Docker base images (\url{https://github.com/gigantum/base-images}),
and users are able to define and configure their own custom images. The
available images include two with R based on Ubuntu Linux have the
\href{https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/}{\texttt{c2d4u}
CRAN PPA} pre-configured for installation of binary R
packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter
in \texttt{r-tidyverse} or both Jupyter \& RStudio in
\texttt{rstudio-server}. The independent image stack can be traced back
to the Gigantum environment and its features. The R images are based on
Gigantum's \texttt{python3-minimal} image, originally to keep the
existing front-end configuration, but also to provide a consistent
Python-to-R interoperability. The \texttt{Dockerfile}s also use build
args to specify bases, for example for different version of NVIDIA CUDA
for GPU
processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.}
so that appropriate GPU drivers can be enabled automatically when
supported. Furthermore, Gigantum's focus lies on environment management
via GUI and ensuring a smooth user interaction, e.g., with reliable and
easy conflict detection and resolution. For this reason, project
repositories store authoritative package information in a separate file
per package, allowing Git to directly detect conflicts and changes. A
Dockerfile is generated from this description that inherits from the
specified base image, and additional custom Docker instructions may be
appended by users, though Gigantum's default base images do not
currently include the \texttt{littler} tool, which is used by Rocker to
install packages within \texttt{Dockerfile}s. Because of these
specifics, instructions from \texttt{rocker/r-ubuntu} could \emph{not}
be readily re-used in this image stack (see
Section~\nameref{conclusions}). Both approaches enable \texttt{apt} as
an installation method, and this is exposed via the GUI-based
environment
management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON
template configuration files, unlike Rocker which relies on plain
\texttt{Dockerfile}s. A minor reason in the inception of the images were
also project constraints requiring a Rocker-incompatible licensing of
the \texttt{Dockerfile}s, i.e., the MIT License.

@@

>>>>

% !TeX root = RJwrapper.tex
\title{The Rockerverse: Packages and Applications for Containerization with R}
\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao, Charlotte Van Petegem}

@@

\abstract{%
The Rocker~Project provides widely-used Docker images for R across
different application scenarios. This articles surveys downstream
projects building upon Rocker and presents the current state of R
packages for managing Docker images and controlling containers. These
use cases and the variety of applications demonstrate the power of
Rocker specifically and containerisation in general. We identified
common themes across this diversity: reproducible environments,
scalability and efficiency, and portability across clouds.
}

@@

The article continues with a brief introduction of containerization
basics and the Rocker~Project, followed by use cases and applications,
starting with the R packages specificly for interacting with Docker,
second-level packages using containers indirectly or only for specific
features, up to complex use cases leveraging containers. We conclude
with a reflection on the landscape of packages and applications and
point out future directions of development.

@@

For statistical computing and analysis centered around R, the
\textbf{Rocker~Project} has provided a variety of Docker containers
since its start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides
several lines of containers spanning to from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for better maintainability (i.e.,
smaller \texttt{Dockerfile}s), composability, and to reduce build time.
Also of note is a series of ``versioned'' containers which match the R
release they contain with the \emph{then-current} set of packages via
the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, who provide findung for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

@@

\hypertarget{image-stacks-for-communities-of-practice}{%
\subsection{Image stacks for communities of
practice}\label{image-stacks-for-communities-of-practice}}

\textbf{Bioconductor} (\url{https://bioconductor.org/}) is an open
source, open development project for the analysis and comprehension of
genomic data \citep{gentleman_bioconductor_2004}. The project consists
of 1823 R software packages as of October 30th 2019, as well as packages
containing annotation or experiment data. \emph{Bioconductor} has a
semi-annual release cycle, each release is associated with a particular
version of R, and Docker images are provided for current and past
versions of \emph{Bioconductor} for convenience and reproducibility. All
images, included are described on the \emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), created with
\texttt{Dockerfile}s maintained on GitHub, and distributed through
Docker~Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
\emph{Bioconductor}'s `base' Docker images are built on top of the
\texttt{rocker/rstudio} image. \emph{Bioconductor} installs packages
based on the R version in combination with the Bioconductor version, and
therefore uses Bioconductor version tagging \texttt{devel} and
\texttt{RELEASE\_X\_Y}, e.g., \texttt{RELEASE\_3\_10}. Past and current
combinations of R and \emph{Bioconductor} will therefore be accessible
via a specific image tags. The \emph{Bioconductor} \texttt{Dockerfile}
selects the desired version of R from Rocker, adds required system
dependencies, and uses the \CRANpkg{BiocManager} package for installing
appropriate versions of \emph{Bioconductor} packages. A strength of this
approach is that the responsibility for complex software configuration
(including customized development) is shifted from the user to the
experienced \emph{Bioconductor} core team. A recent audit of the
\emph{Bioconductor} image stack \texttt{Dockerfile} led to the
deprecation of several community-maintained images, because the numerous
specific images became too hard to understand, complex to maintain, and
cumbersome to extent. As part of the simplification, a recent innovation
is to produce a \texttt{bioconductor\_docker:devel} image to emulate the
\emph{Bioconductor} nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the
\emph{system dependencies} needed to install and check almost all (1813
out of 1823) \emph{Bioconductor} software packages and saves users and
package developers from managing these themselves. Furthermore the image
is configured so that \texttt{.libPaths()} has
\texttt{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can persist
installed packages across Docker sessions or image updates. Many R users
pursue flexible work flows tailored to particular analysis needs, rather
than standardized work flows. The new \texttt{bioconductor\_docker}
image is well-suited for this pattern, while
\texttt{bioconductor\_docker:devel} provides developers with a test
environment close to \emph{Bioconductor}'s build system.

\label{datascience} \textbf{Data Science} is a widely discussed topic
among all academic disciplines \citep[e.g.,][]{donoho_50_2017}. The
discussions shed a light on the tools and craftspersonship behind the
analysis of data with computational methods. The practice of Data
Science often involves a combination tools and software stacks and
requires a cross-cutting skillset. This complexity and an inherent
concern for openness and reproducibility in the Data Science community
lead to Docker being used widely. The remaineder of this section
presents exemplary Docker images and image stacks featuring R intended
for Data Science. The
\href{https://github.com/jupyter/docker-stacks/}{\emph{Jupyter Docker
Stacks}} project are a set of ready-to-run Docker images containing
